---
title: "DS2020_Final_Project"
author: "Trey Hartung"
date: "2026-05-12"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Analysis on the Effects of Gaming on Academic Performance
#### Trey Hartung

## Introduction
The purpose for this project is to figure out the relationship between gaming and academic performance for students. Many students students play video games, myself included, so it is important to understand how exactly this can affect our studies.

For this purpose I ask the following questions:

1. Does gaming time affect student grades?

2. Do study hours improve academic performance?

3. How do gaming time and study hours affect stress level?

These questions are the main things I am curious about and will try to answer throughout this project. With the answers to these questions I will be able to find meaningful conclusions on how gaming cam relate to academic performance.

## Data
### Structure
The dataset can be found at https://www.kaggle.com/datasets/aiexplorer77/gaming-vs-academic-performance . This website provides the csv file containing the data that will be used for this project. The dataset is composed of 8000 observations with 14 variables each. For this project the most instrumental variables for my questions will be ones such as gaming hours, study hours, stress level, and grades. The data is already in an easy to work with form, so no cleaning will be necessary after loading the dataset.

### Variables
* student_id: The unique identifier for each student.
* age: The age of the student.
* gender: The gender of the student.
* gaming_hours: The average number of hours spent gaming each day.
* study_hours: The average number of hours spent studying each day.
* sleep_hours: The average number of hours spent sleeping each day.
* attendance: The percentage of classes that the student attends.
* gaming_genre: The student's preferred genre of game.
* social_activity: The number of hours spent on social activities each day.
* device_usage: The total daily screen time each day.
* reaction_time_ms: The cognitive reaction time in milliseconds.
* addiction_score: The Estimated gaming addiction level.
* stress_level: The Stress category of the student.
* grades: The academic performance score of the student.

## Results
```{r}
library(readr)
library(tidyverse)

student_data <- read_csv("Gaming_Academic_Performance.csv")
```

### Does gaming time affect student grades?

#### Correlation
```{r}
cor(student_data$gaming_hours, student_data$grades)
```
This correlation value means that gaming hours and grades have a moderate negative linear relationship.Based on this, we can expect a graph of the data to show grades get worse the more hours spent gaming a student has.

#### Scatterplot
```{r}
scatterplot1 <- ggplot(student_data, aes(x = gaming_hours, y = grades)) +
  geom_point(alpha = 0.4, color = "blue") +
  geom_smooth(method = "lm", color = "red") +
  labs(
    title = "Gaming Hours vs Grades",
    x = "Gaming Hours",
    y = "Grades"
  )

print(scatterplot1)
```

The plot above shows exactly what was expected from the correlation value, a mostly linear, negative relationship. It shows that students who spend more time gaming generally have lower grades than those who do not game as much.

#### Linear regression
```{r}
LRmodel1 <- lm(grades ~ gaming_hours, data = student_data)
summary(LRmodel1)
```
I chose to do a linear regression model with this data to see just how much gaming lowers grades. The number I focus on from the summary data is the estimate for gaming hours in the "Coefficients" section. This tells us that with every hour spent gaming, grades get worse by 5.35 points on average.

### Do study hours improve academic performance?

#### Correlation
```{r}
cor(student_data$study_hours, student_data$grades)
```
This value tells us that study hours and grades have a strong positive relationship. So we expect grades to increase alongside hours spent studying.

#### Scatterplot
```{r}
scatterplot2 <- ggplot(student_data, aes(x = study_hours, y = grades)) +
  geom_point(alpha = 0.4, color = "darkgreen") +
  geom_smooth(method = "lm", color = "black") +
  labs(
    title = "Study Hours vs Grades",
    x = "Study Hours",
    y = "Grades"
  )

print(scatterplot2)
```

This plot reaffirms what the correlation value tells us. There is a clear trend that students who studied for more hours had higher scores on average than those who did not study as much.

#### Linear regression
```{r}
LRmodel2 <- lm(grades ~ study_hours, data = student_data)
summary(LRmodel2)
```
This linear regression model tells us that for every hour studied, grades will increase by 6.38 on average. 

### How do gaming time and study hours affect stress level?

#### Correlations
```{r}
student_data$stress_numeric <- ifelse(student_data$stress_level == "Low", 1, ifelse(student_data$stress_level == "Medium", 2, 3))

cor(student_data$gaming_hours, student_data$stress_numeric)
cor(student_data$study_hours, student_data$stress_numeric)
```
For this question, the first thing I did was convert the stress levels into numeric values to make them easier to use in statistical analysis. These correlation values indicate that gaming hours and study hours have moderate negative and weak positve relationships with stress respectively.

#### Multiple regression
```{r}
stress_model <- lm(stress_numeric ~ gaming_hours + study_hours, data = student_data)
summary(stress_model)
```
Based on this regression model, we see that on average, more time spent gaming reduces the level of stress that students feel, whereas studying does the opposite. It is important to note, however, that stress level was originally categorical in nature. This means that analyzing them numerically is not always appropriate. In this case it is still useful to at least see the direction of the relationship (positive vs negative) for gaming and studying on stress.

#### Gaming vs stress
```{r}
stress_plot1 <- ggplot(student_data,
                aes(x = gaming_hours, y = stress_numeric)) +
  geom_jitter(alpha = 0.3, color = "purple") +
  geom_smooth(method = "lm", color = "black") +
  labs(
    title = "Gaming Hours vs Stress Level",
    x = "Gaming Hours",
    y = "Stress Level"
  )

print(stress_plot1)
```

This plot for gaming hours and stress level shows that for any number of hours spent gaming, some students always had high levels of stress. It also shows that at 5+ hours of gaming, there is no more medium level stress reported and a large amount of low level stress reported. This seems somewhat unnatural so it would be proper to investigate how this data was sampled to look for an explanation for this. 

#### Study hours vs stress
```{r}
stress_plot2 <- ggplot(student_data,
                aes(x = study_hours, y = stress_numeric)) +
  geom_jitter(alpha = 0.3, color = "orange") +
  geom_smooth(method = "lm", color = "black") +
  labs(
    title = "Study Hours vs Stress Level",
    x = "Study Hours",
    y = "Stress Level"
  )

print(stress_plot2)
```

A similar plot is used for modeling study hours and stress level. This plot shows that at all amounts of hours spent studying, there are always students reporting low and medium levels of stress. Below 7 hours of studying, there are no reports of high stress levels, whereas at 7+ hours of studying, there are frequent reports of high level stress. This also seems somewhat unnatural that there is no high level stress until after 7 hours of studying. Further investigation to confirm this data would be appropriate.

## Conclusion

In conclusion, my analysis on the effects of gaming on academic performance confirm the intuitive answers to how gaming and studying relate to grades and stress. However, putting numerical values to these effects allow for better planning and/or management for how students spend their time and what they can expect the results of their choices to be. For instance, based on the correlation and coefficient values, it is shown that studying has a greater affect on grade positively than gaming does negatively. Similarly, gaming is more effective at lowering stress than studying is as raising stress. This suggests that a balance of gaming and studying is important in order to keep grades high without stressing yourself out too much.

Further analysis would be appropriate to get even more information on this topic. factoring in aspects like how much sleep students get or how often they attend classes would increase the understanding on what exactly affects grades and stress the most. It is also important to consider that lower grades that appear to come from many hours of gaming could really be caused simply by low hours studying. 

Overall, this project was worthwhile to me as a student who enjoys video games while also understanding the importance of studying. I will certainly think back to this project whenever I debate with myself on how to best spend my time.